Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-29 parallel-tractability-J-v9/introduction.tex
splitting the main file for better organization

%\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
 \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
\usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}

\usepackage{amsmath}

\usepackage{CJK}
\usepackage{color}

\usepackage{enumitem}
\usepackage{rotating}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

%\journal{Nuclear Physics B}

\hyphenation{data-log}

\begin{document}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{Parallel Tractability of Ontology\\
Materialization: Technique and Practice}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}

\author[add1]{Zhangquan Zhou \corref{cor1}}
\ead{quanzz1129@gmail.com}
\cortext[cor1]{Corresponding author}

\author[add1]{Guilin Qi}
\ead{gqi@seu.edu.cn}

\author[add2]{Birte Glimm}
\ead{birte.glimm@uni-ulm.de}

\address[add1]{
    School of Computer Science, Southeast University, China
}

\address[add2]{
    Institute of Artificial Intelligence, University of Ulm
}

\begin{abstract}
  Materialization is an important reasoning service for many
  ontology-based applications, but the rapid growth of semantic data
  poses the challenge to efficiently perform materialization on
  large-scale ontologies. Parallel materialization algorithms work
  well for some ontologies, which is not always explained by the
  theoretical results. For example, for the YAGO ontology
  parallelization works well, although the reasoning problem for the
  used ontology language is not in \texttt{NC}, i.e.\ the theoretical
  complexity class for \emph{parallel tractability}. This motivates us
  to study the problem of \emph{parallel tractability} of ontology
  materialization from a theoretical perspective. We focus on the
  datalog rewritable ontology languages DL-Lite and Description Horn
  Logic (DHL) and propose algorithms, called \texttt{NC} algorithms,
  to identify classes of ontologies for which materialization is
  tractable in parallel. To verify the practical usability of the
  above results, we analyze real-world datasets and show that for many
  ontologies expressed in DHL materialization is tractable in
  parallel. The implementation of our optimized parallel algorithm
  shows performance improvements over the highly optimized
  state-of-the-art reasoner RDFox on ontologies for which
  materialization is tractable in parallel.
\end{abstract}

\begin{keyword}
ontology \sep materialization \sep datalog \sep parallel tractability \sep NC complexity
\end{keyword}

\end{frontmatter}

%% \linenumbers

@@

\section{Background Knowledge}
\label{sec:background}

In this section, we introduce some basic notions that are needed to introduce our approach.

\subsection{Datalog}

We discuss the main issues in this paper using standard datalog notions.
In datalog \cite{database}, a \emph{term} is a variable or a constant. An \emph{atom} $A$
is defined by $A\equiv p(t_1,...,t_n)$ where $p$ is a \emph{predicate} (or \emph{relational})
name, $t_1,...,t_n$ are terms, and $n$ is the arity of $p$. If all the terms in an atom $A$ are
constants, then $A$ is called a \emph{ground atom}.
A datalog \emph{rule} is of the form: `$B_1,...,B_n\rightarrow H$',\footnote{In datalog rules, a comma
represents a Boolean conjunction `$\wedge$'.} where $H$ is referred to as
the \emph{head atom} and $B_1,...,B_n$ the \emph{body atoms}. Each variable in the head atom
of a rule must occur in at least one body atom of the same rule. A \emph{fact} is a rule of
the form `$\rightarrow H$', i.e., a rule with an empty body and the head $H$ being a ground atom.
A datalog program $P$ consists of rules and facts.
A \emph{substitution} $\theta$ is a partial mapping of variables to constants.
For an atom $A$, $A\theta$ is the result of replacing each variable $x$ in $A$
with $\theta(x)$ if the latter is defined. We call $\theta$ a \emph{ground substitution}
if each defined $A\theta$ is a ground atom.
A \emph{ground instantiation} of a rule is obtained by applying a ground substitution on all
the terms in this rule with respect to a finite set of constants occurring in $P$.
Furthermore the ground instantiation of $P$, denoted by $P^*$,
consists of all ground instantiations of rules in $P$.
The predicates occurring only in the body  of some rules are called \emph{EDB predicates},
while the predicates that may occur as head atoms are called \emph{IDB predicates}.

\subsection{DHL and DL-Lite}

In what follows, we use $\textbf{CN}$, $\textbf{RN}$ and $\textbf{IN}$
to denote three disjoint countably
infinite sets of \emph{concept names, role names}, and \emph{individual names} respectively.
The set of roles is defined as $\textbf{R}:=\textbf{RN}\cup\{R^-|R\in\textbf{RN}\}$
where $R^-$ is the \emph{inverse role} of $R$.
For ease of discussion, we focus on the \emph{simple forms} of axioms shown
in the left column of Table~1. These simple forms can be obtained by using
well-known \emph{structure transformation} techniques \cite{KrotzschRH07,Kazakov09}.

DHL (short for \emph{description horn logic}) \cite{GrosofHVD03} is introduced as an
intersection of description logic (DL) and datalog in terms of expressivity.
We define a DHL ontology $\mathcal{O}$ as a triple:
$\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$, where
$\mathcal{T}$ denotes the TBox containing axioms of the forms (T1-T3);
$\mathcal{R}$ is the RBox that is a set of axioms of the forms (R1-R3);
$\mathcal{A}$ is the ABox containing \emph{assertions} of the forms (A1) and (A2).
In an axiom of either of  the forms (T1-T3 and R1-R3), concepts $A_{(i)}$ and $B$ are either
concept names, \emph{top concept} ($\top$) or \emph{bottom concept} ($\bot$); $R$ and $S_{(i)}$
are roles in $\textbf{R}$.
For an axiom of the form $A\sqsubseteq\forall R.B$ that is also allowed in DHL, we only consider its
equivalent form $\exists R^-.A\sqsubseteq B$.

DHL is related with other ontology languages.
First, DHL is essentially a fragment of the description logic Horn-$\mathcal{SHOIQ}$ with
disallowing \emph{nominal}, \emph{number restriction} and
right-hand \emph{existential restriction} ($A\sqsubseteq\exists R.B$).
Second, the expressivity of DHL covers that of RDFS to some extent \cite{GrosofHVD03}.
Reasoning with RDFS ontologies is \texttt{NP}-complete \cite{Horst05}
and, thus, is not parallelly tractable.
However, by applying some simplifications and restrictions, RDFS ontologies can be
expressed in DHL \cite{GrosofHVD03} that has \texttt{PTime}-complete complexity for
materialization.

\begin{table}
\begin{center}
\begin{tabular}{lcc}
\multicolumn{3}{c}{\textbf{Table 1: Axioms and corresponding datalog rules}}\\
\hline
&~~~~~~~~~~~~~~~~~~~~~~~~Axioms~~~~~~~~~~~~~~~~~~~~~~~~&~~~~~~~~~~~~~~~~~~~~~~~~Datalog Rules~~~~~~~~~~~~~~~~~~~~~~~~\\
\hline
\hline
(T1)& $A\sqsubseteq B$& $A(x)\rightarrow B(x)$\\
(T2)& $A_1\sqcap A_2\sqsubseteq B$& $A_1(x), A_2(x)\rightarrow B(x)$\\
(T3)& $\exists R.A\sqsubseteq B$& $R(x,y), A(y)\rightarrow B(x)$\\
(T4)& $A\sqsubseteq\exists R$& $A(x)\rightarrow R(x, o_{R}^A)$\\

\hline
(R1)& $S\sqsubseteq R$& $S(x,y)\rightarrow R(x,y)$\\
(R2)& $S\sqsubseteq R^-$& $S(x,y)\rightarrow R(y,x)$\\
(R3)& $R\circ R\sqsubseteq R$& $R(x,y), R(y,z)\rightarrow R(x,z)$\\
(R4)& $R_1\circ R_2\sqsubseteq R$&$R_1(x,y),R_2(y,z)\rightarrow R(x,z)$\\

\hline
(A1)& $A(a)$& $A(a)$\\
(A2)& $R(a,b)$& $R(a,b)$\\
\hline
\end{tabular}
\label{tab:dhl}
\end{center}
\end{table}

In the initial work of DHL \cite{GrosofHVD03}, \emph{complex role inclusion axioms} (complex RIAs) of
the form $R_1\circ...\circ R_n\sqsubseteq R$ are not considered, although they can be
naturally transformed to datalog rules.
In this paper, we also consider an extension of DHL (denoted by DHL($\circ$))
that allows complex RIAs. Since a complex RIA can be transformed to
several axioms of the form (R4), we then require that an
RBox $\mathcal{R}$ of a DHL($\circ$) ontology can contain
axioms of the forms (R1-R4). Note that (R3) is actually a special
case of (R4).

DL-Lite is a group of ontology languages designed for highly-efficient
query answering through knowledge
and underpins the OWL profile OWL QL \cite{CalvaneseGLLR07}.
DL-Lite$_{core}$ and DL-Lite$_{\mathcal{R}}$ are the two basic fragments
in DL-Lite.
DL-Lite$_{core}$ requires that an TBox contains axioms of the forms (T1) $A\sqsubseteq B$,
(T2) $A_1\sqcap A_2\sqsubseteq B$ where $B\equiv\bot$, (T3) $\exists R.A\sqsubseteq B$
where $A\equiv\top$ or $B\equiv\bot$, and (T4) $A\sqsubseteq\exists R$,
an ABox contains \emph{assertions} of the forms (A1) and (A2), and
RBox is not involved. DL-Lite$_{\mathcal{R}}$ is an extension of DL-Lite$_{core}$,
which allows involving RBoxes that contain axioms of the forms (R1-R2).
If not specially specified, \emph{a DL-Lite ontology denotes a DL-Lite$_{\mathcal{R}}$ ontology
in the following paragraphs}.

\subsection{Ontology Materialization via Datalog Programs}

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ can be transformed to a datalog program
(see the corresponding rules in the right column of Table~1).
Note that, for an axiom of the form (T4) $A\sqsubseteq\exists R$, it should be transformed to
a first-order logic rule `$A(x)\rightarrow \exists y(R(x,y))$' where $y$ is called a \emph{free variable} (it is
not occurring in the head atom $A(x)$).
In this work, such a free variable $y$ is eliminated via Skolemisation into a new individual $o_{R}^A$
that corresponds to concept $A$ and role $R$. One can check that the elimination of free variables
does not influence the results of materialization by according to \cite{CalvaneseGLLR07}.

In what follows, for an ontology $\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$,
we also use $P=\langle R, \textbf{I}\rangle$ to represent the corresponding datalog program
where $R$ is the set of rules transformed from the axioms in $\mathcal{T}$ and $\mathcal{R}$,
$\textbf{I}$ is the set of facts that are directly copied from the assertions in $\mathcal{A}$.
Further, we use $R_1\sqsubseteq_{*}R_2$ to denote the smallest transitive reflexive relation
between roles such that $R_1\sqsubseteq R_2\in\mathcal{R}$ implies $R_1\sqsubseteq_{*}R_2$
and $R_1^-\sqsubseteq_{*}R_2^-$. In this paper, we also use the
notion of \emph{simple role}, which is initially proposed to restrict the
usage of highly expressive ontology languages \cite{HorrocksS04}.
Specifically, a role $S\in\textbf{R}$ is \emph{simple} if, (1) it has no subrole (including $S$)
occurring on the right-hand side of axioms of the forms (R3) and (R4); (2) $S^-$ is simple.

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,...,B_n\rightarrow H\in R, B_i\theta\in\textbf{I} (1\leq i\leq n)\}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}
for any class of datalog programs.

\subsection{The NC Complexity}

The parallel complexity class NC, known as Nick$'$s
Class \cite{Raymond95}, is studied by theorists as a parallel complexity class
where each decision problem can be efficiently solved in parallel.
Specifically, a decision problem in the NC class
can be solved in poly-logarithmic time on a PRAM (parallel, random-access machine) with
a polynomial number of processors. We also say that an NC problem can be solved
in \emph{parallel poly-logarithmic time}.
Although the NC complexity is a theoretical analysis tool,
it has been shown that many NC problems can be solved efficiently in practice \cite{Raymond95}.

From the perspective of implementations, the NC problems are also highly
parallel feasible for other parallel models like BSP \cite{Valiant90}
and MapReduce \cite{KarloffSV10}. The NC complexity is originally defined
as a class of decision problems. Since we study the problem of materialization, we do not
require in this work that a problem should be a decision problem in NC.
In addition, since many parallel reasoning systems (see related work in Section~\ref{sec:related})
are implemented on shared-memory platforms, we
study all the issues in this work by assuming that the running machines are in
shared-memory configurations.

\section{Parallel Tractability of Datalog Programs}
\label{sec:ptclass}

Our target is to find for which kinds of ontologies (not ontology languages)
materialization is parallelly tractable.
Since we assume that for any class of datalog programs where the rule
set of each datalog program is fixed, the materialization problem is thus in data complexity
$\texttt{PTime}$-complete, which is considered to be inherently sequential in the worst
case \cite{Raymond95}. In other words, the materialization problem on general
datalog programs cannot be solved in parallel poly-logarithmic time unless P=NC.
Thus, we say that \emph{materialization on a class of datalog programs is parallelly tractable
if there exists an algorithm that handles this class of datalog programs and runs in parallel
poly-logarithmic time (this algorithm is also called an NC algorithm)}.
In this section, we identify such classes of datalog programs by studying different NC algorithms.

\subsection{Parallel Tractable Classes}

We first give the following definition for a parallel tractable class of datalog programs
where an NC algorithm exists for handling each datalog program in this class.

\begin{definition}\label{def:ptd}
\textbf{(Parallelly Tractable Class)} Given a class $\mathcal{D}$ of datalog programs where the
same rule set is shared, we say that $\mathcal{D}$ is a parallelly tractable datalog program (\texttt{PTD}) class
if there exists an NC algorithm that performs sound and complete materialization for each datalog program
in $\mathcal{D}$. The corresponding class of ontologies of $\mathcal{D}$ is called a
parallelly tractable ontology (\texttt{PTO}) class.
\end{definition}

According to the above definition, if we find an NC algorithm $\mathsf{A}$
for datalog materialization, then we can identify a \texttt{PTD} class $\mathcal{D}_{\mathsf{A}}$,
which is the class of all datalog programs that can be handled by $\mathsf{A}$.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not NC algorithms
due to their \texttt{PTime}-complete complexity, since they are designed for handling general datalog programs.
Thus we give our NC algorithms.
In the following, we first give a parallel materialization algorithm that works for
general datalog programs. We then restrict this algorithm to an NC version
and identify the target \texttt{PTD} class.

\subsection{Materialization Graph}

In order to give a parallel materialization algorithm,
we introduce the notion of \emph{materialization graph}.
It makes the analysis of the given algorithm convenient.

\begin{definition}
\textbf{(Materialization Graph)}\label{def:mg}
A \emph{materialization graph}, with respect to
a datalog program $P=\langle R, \textbf{I}\rangle$, is a directed acyclic graph
denoted by $\mathcal{G}=\langle V, E\rangle$ where,
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $V$ is the node set and $V\subseteq T_R^{\omega}(\textbf{I})$;
  \item $E$ is the edge set and $E\subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$.
\end{enumerate}
For each edge of the form $e(v_1, v_2)$ where $v_1$ and $v_2$ are the two nodes,
we say that node $v_1$ is the parent (node) of node $v_2$ and node $v_2$ is the child (node)
of node $v_1$. Further, $\forall H,B_1,...,B_n\in V$, $\mathcal{G}$ satisfies the following conditions:
\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
  \item $H$ has a parent node or a child node;
  \item $H$ is the original fact of $P$ if $H$ has the in-degree of $0$;
  \item $B_1,...,B_n\rightarrow H\in P^*$ is satisfied if $e(B_1, H),...,e(B_n, H)\in E$ and
  $B_1,...,B_n$ are \d{all} the parents of $H$.
\end{enumerate}
\end{definition}

For some derived atom $H$, there may exist several rule instantiations where $H$
occurs as a head atom.
This also means that $H$ can be derived in different ways.
The condition in the definition above results in only one way of deriving $H$ being
described by a materialization graph.
Suppose $\mathcal{G}$ is a materialization graph, the nodes whose in-degree is 0 are the
original facts in $\textbf{I}$.
The size of $\mathcal{G}$, denoted by $|\mathcal{G}|$, is the number of nodes in $\mathcal{G}$.
The depth of $\mathcal{G}$, denoted by \texttt{depth}($\mathcal{G}$), is the maximal length of a path
in $\mathcal{G}$.
We next give an example of a materialization graph.\\

\begin{example}\label{exp:mg}
Consider a DHL($\circ$) ontology $\mathcal{O}_{ex_1}$ where the TBox is
$\{\exists R.A\sqsubseteq A\}$, the RBox is $\{S\circ R\sqsubseteq R\}$ and
the ABox is $\{A(b),R(a_1,b),S(a_i,a_{i-1})\}$ for $2\leq i\leq k$ and
$k$ is an integer greater than $2$.
The corresponding datalog program of this ontology is $P_{ex_1}=\langle R, \textbf{I}\rangle$
where $\textbf{I}$ contains all the assertions in the ABox
and $R$ contains the two rules `$R(x,y),A(y)\rightarrow A(x)$'
and `$S(x,y),R(y,z)\rightarrow R(x,z)$'.
The graph in Figure~\ref{fig:mg} is a materialization graph with respect to $P_{ex_1}$,
denoted by $\mathcal{G}_{ex_1}$.
The nodes whose in-degree is 0 are the original facts in $\textbf{I}$;
each of the other nodes corresponds to a ground instantiation of some rule.
For example, the node $A(a_k)$ corresponds to the ground rule instantiation
`$R(a_k,b),A(b)\rightarrow A(a_k)$'.
The size of this materialization graph is the number of nodes, that is $3k$.
The depth of $\mathcal{G}_{ex_1}$ is $k$.
\end{example}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-mg.eps}
\caption{An example of a materialization graph.}
\label{fig:mg}
\end{center}
\end{figure}

We say that a materialization graph $\mathcal{G}$ is a \emph{complete materialization graph}
when $\mathcal{G}$ contains all ground atoms in $T_R^{\omega}(\textbf{I})$.
The set of nodes in a complete materialization graph is actually the result
of materialization.
Thus, the procedure of materialization can be transformed to the construction of
a complete materialization graph.
We pay our attention to complete materialization graphs and do not distinguish it to the
notion `materialization graph'.
It should also be noted that there may exist several materialization graphs for a datalog program.

\subsection{A Basic Parallel Algorithm}
\label{sec:alg-bsc}

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{bsc}$) that constructs a materialization graph for a given datalog program.
Our computational model is a PRAM (parallel, random-access machine) that is mostly used to analyze the parallel
complexity. This model allows us giving \emph{the parallel assumption}:
for any datalog program $P=\langle R, \textbf{I}\rangle$,
any substitution of some atom and any rule instance in $P^*$ can be mapped to a unique memory
location; further, a one-to-one relation can be established between processors and rule instances.
Under this assumption, a processor can check the applicability of its corresponding
rule instance and access the state of an atom occurring in this rule instance in constant time.
Algorithm~$\mathsf{A}_{bsc}$ is then given as follows.\\

\noindent\texttt{Algorithm~$\mathsf{A}_{bsc}$}. Given a datalog program $P=\langle R, \textbf{I}\rangle$,
the algorithm returns a materialization graph $\mathcal{G}$ of $P$.
Suppose we have $|P^*|$ processors, and each rule instantiation in $P^*$ is
assigned to one processor.
Initially $\mathcal{G}$ is empty. The following three steps are then performed:
\begin{enumerate}[leftmargin=8ex,label=(\textit{Step \arabic*}),ref=Step~\arabic*]
\item Add all facts in $\textbf{I}$ to $\mathcal{G}$.\label{alg1:addFacts}
\item For each rule instantiation $B_1,...,B_n\rightarrow H$, if the body atoms are all
    in $\mathcal{G}$ while $H$ is not in $\mathcal{G}$,
    the corresponding processor adds $H$ to $\mathcal{G}$ and creates edges pointing
    from $B_1,...,B_n$ to $H$.\label{alg1:updateG}
\item If no processor can add more nodes and edges to $\mathcal{G}$, terminate, otherwise iterate \ref{alg1:updateG}.\label{alg1:halt}\hfill$\Box$
\end{enumerate}

\begin{example}
We consider the datalog program $P_{ex_1}$ in Example~\ref{exp:mg} again,
and perform Algorithm~$\mathsf{A}_{bsc}$ on it.
Initially, all the facts ($A(b),R(a_1,b),S(a_2,a_1),...,S(a_{k},a_{k-1})$) are added to the
result $\mathcal{G}_{ex_1}$ (\ref{alg1:addFacts}).
Then in different iterations of \ref{alg1:updateG}, the remaining nodes are added to
$\mathcal{G}_{ex_1}$ by different processors.
For example a processor $p$ is allocated a rule instantiation `$R(a_2,b),A(b)\rightarrow A(a_2)$'.
Then, processor $p$ adds $A(a_2)$ to $\mathcal{G}_{ex_1}$ after it checks that
$A(b)$ and $R(a_2,b)$ are in $\mathcal{G}_{ex_1}$.
Algorithm~$\mathsf{A}_{bsc}$ halts when $A(a_k)$ has been added to $\mathcal{G}_{ex_1}$ (\ref{alg1:halt}).
\end{example}

Lemma~\ref{lemma:a1} shows the correctness of Algorithm~$\mathsf{A}_{bsc}$ and that, for any datalog program $P$,
Algorithm~$\mathsf{A}_{bsc}$ always constructs a materialization graph with the minimum depth among all the
materialization graphs of $P$. The detailed proofs of Lemma~\ref{lemma:a1} and other lemmas and theorems can be found
in the appendix.

\begin{lemma}
\label{lemma:a1}
Given a datalog program $P=\langle R, \textbf{I}\rangle$, we have
\begin{enumerate}[leftmargin=4ex]
\item Algorithm~$\mathsf{A}_{bsc}$ halts and returns a materialization graph $\mathcal{G}$ of $P$;
\item $\mathcal{G}$ has the the minimum depth among all the materialization graphs of $P$.
\end{enumerate}
\end{lemma}

\begin{proof}[Proof sketch] This lemma can be proved by performing
an induction on $T_R^{\omega}(\textbf{I})$.
The stage (see the related contents in Section~\ref{sec:background}) of $P$
is the lower-bound of the depth of the materialization graphs. Based on the previous induction,
one can further check that, for the materialization graph $\mathcal{G}$ constructed by Algorithm~$\mathsf{A}_{bsc}$,
its depth equals the depth of the stage.
\end{proof}

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{bsc}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any EDB predicate, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instances in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instances is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{bsc}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus,
the whole computing time of Algorithm~$\mathsf{A}_{bsc}$ turns out to be $c_1+l\cdot c_2$
where $l$ denotes the number of iterations of \ref{alg1:updateG}.

When we say that an algorithm is an NC algorithm. It should meet two requirements: first,
it works on a polynomial number of processors; second, it halts in poly-logarithmic time.
As discussed above, Algorithm~$\mathsf{A}_{bsc}$ meets the first requirement.
If we want to restrict Algorithm~$\mathsf{A}_{bsc}$ to be an NC algorithm,
we can make the number of iterations of \ref{alg1:updateG} to be poly-logarithmically bounded.
We use the symbol $\psi$ to denote a poly-logarithmically bounded function.
For any datalog program, if we restrict that the number of iterations of \ref{alg1:updateG} is bounded by $\psi(|\textbf{I}|)$,
the computing time of Algorithm~$\mathsf{A}_{bsc}$
is $c_1+\psi(|\textbf{I}|)\cdot c_2$. With this restriction, Algorithm~$\mathsf{A}_{bsc}$ is an NC algorithm denoted by
$\mathsf{A}_{bsc}^{\psi}$.

Based on $\mathsf{A}_{bsc}^{\psi}$, we can identify a class of datalog programs
$\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ such that all the datalog programs in it can be handled
by $\mathsf{A}_{bsc}^{\psi}$.
It is obvious that $\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ is a \texttt{PTD} class.
We use the following theorem to further show that this class can be captured based on
the materialization graphs of the datalog programs in $\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$.

\begin{theorem}\label{theorem:a1}
For any datalog program $P$, $P\in\mathcal{D}_{\mathsf{A}_{bsc}^{\psi}}$ iff $P$ has a
materialization graph whose depth is upper-bounded by $\psi(|P|)$.
\end{theorem}

\begin{proof}[Proof sketch]
We can first prove that the number of iterations
of \ref{alg1:updateG} is actually the depth of the constructed materialization
graph. This theorem then follows by considering
Lemma~\ref{lemma:a1}.
\end{proof}

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{ex_1}$, where the rule set
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the fact
set varies according to $k$. The algorithm $\mathsf{A}_{bsc}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{ex_1}$.
It can be checked that, for some materialization graph $\mathcal{G}_{ex_1}$
corresponding to a datalog program in $\mathbb{P}_{ex_1}$, \texttt{depth}($\mathcal{G}_{ex_1}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{ex_1}$ are polynomial of $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{bsc}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{ex_1}$.
However there indeed exists an NC algorithm that can handle $\mathbb{P}_{ex_1}$.
We discuss this in the next part.

\subsection{Optimizing Algorithm~$\mathsf{A}_{bsc}$ via the Single-Way Derivability}

In this part, we optimize Algorithm~$\mathsf{A}_{bsc}$ such that $P_{ex_1}$
can be handled. Based on the optimized variant of Algorithm~$\mathsf{A}_{bsc}$,
we can identify another \texttt{PTD} class.

We discuss our optimization based on a specific case in Example~\ref{exp:general}.
We find that, in this kind of case, the construction of a materialization graph can be accelerated.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig-general.eps}
\caption{A partial materialization graph.}
\label{fig:general}
\end{center}
\end{figure}

\begin{example}\label{exp:general}
Consider a snapshot of Algorithm~$\mathsf{A}_{bsc}$ in
Figure~\ref{fig:general}. A materialization graph $\mathcal{G}$
is being constructed for some datalog program $\langle R, \textbf{I}\rangle$.
The nodes in the dashed box denote the ones that have been added to $\mathcal{G}$.
In this snapshot, $v_0$ has been newly added
to $\mathcal{G}$ in the $l^{th}$ ($l\geq 1$) iteration.
Each of the nodes $v_i$ ($1\leq i\leq k$) has almost one parent node being not in $\mathcal{G}$,
while $v'$ has two parent nodes being not in $\mathcal{G}$.
All of the nodes $v_i$ ($1\leq i\leq k$) and $v'$ would be added to $\mathcal{G}$
afterwards.
\end{example}

In Example~\ref{exp:general}, $v_k$ would be added to $\mathcal{G}$ after \emph{at least} $k$
iterations by performing Algorithm~$\mathsf{A}_{bsc}$.
We can check that each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) in this stage follows the condition:
\emph{if the parent node $v_{i-1}$ is derivable, the child node $v_i$ has to be derivable} (this
is because that, for node $v_i$, node $v_{i-1}$ is the only parent node that has not been added to $\mathcal{G}$).
We call this condition \emph{the single-way derivability condition}, which intuitively says that the derivability of a
node only depends on one of its parent nodes.
Observe that node $v_k$ is reachable from node $v_0$ through the path $\tau=(v_0,v_1,...,v_k)$ where
each edge $(v_{i-1},v_i)$ ($1\leq i\leq k$) follows the single-way derivability condition;
we call such a path a \emph{single-way derivable} (SWD) path.
Further, the starting node $v_0$ in $\tau$ has been added to $\mathcal{G}$; in other words,
$v_0$ is derivable. Thus, all of the nodes $v_i$ ($1\leq i\leq k$) can be added to
$\mathcal{G}$ immediately.
Based on the above idea, we optimize Algorithm~$\mathsf{A}_{bsc}$ by checking whether such an SWD path
exits for some node $v$; if there is an SWD path for node $v$, this node can be added to $\mathcal{G}$
due to the single-way derivability condition.

Backing to Example~\ref{exp:general}, since
there exists an SWD path $(v_0,...,v_i)$ for each node $v_i$ ($1\leq i\leq k$), all of these nodes
can be added to $\mathcal{G}$ right after $v_0$. On the other hand,
node $v'$ has no SWD path, since it has two parent nodes being not in $\mathcal{G}$.

We next discuss how to determine the existence of SWD paths. Note that, SWD paths require us to
describe the reachability between two nodes. To this extent, we use a
binary transitive relation $\texttt{rch} \subseteq T_R^{\omega}(\textbf{I})\times T_R^{\omega}(\textbf{I})$,
e.g., \texttt{rch}$(v_1,v_2)$ means that $v_2$ is reachable from $v_1$.
In each iteration of \ref{alg1:updateG} in Algorithm~$\mathsf{A}_{bsc}$, we further compute a \texttt{rch}
relation (denoted by $S_{\textit{\tiny rch}}$) by performing the following process:

\begin{description}[leftmargin=2ex]
\item[(\textbf{\dag})] \emph{For each rule instantiation of the form $B_1,..,B_i,..,B_n\rightarrow H$
where $H$ has not been added to $\mathcal{G}$}:
\begin{enumerate}[leftmargin=2ex]
\item \emph{if the body atoms $B_1,...,B_n$ have all been added to $\mathcal{G}$, put \texttt{rch}$(B_1,H),...,$ \texttt{rch}$(B_n,H)$ in $S_{\textit{\tiny rch}}$};
\item \emph{if $B_i$ is the only node in the body that has not been added to $\mathcal{G}$,
    put \texttt{rch}$(B_i,H)$ in $S_{\textit{\tiny rch}}$}.\hfill$\Box$
\end{enumerate}
\end{description}

We then compute the transitive closure (denoted by $S^*_{\textit{\tiny rch}}$) with respect to
$S_{\textit{\tiny rch}}$. Based on the transitive closure, we can perform the optimization that,
for a node $v$, if there is a relation $\texttt{rch}(v',v)\in S^*_{\textit{\tiny rch}}$
where $v'$ has been added to $\mathcal{G}$, $v$ has an SWD path and can be added to $\mathcal{G}$.
The following algorithm applies this optimization strategy.\\

>>>>

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "parallel-tractability-J"
%%% End:
